Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2023-07-26 docs/ms-draft.Rmd
coauthors ms review

---
title: "Probabilistic bioregion models suggest seabird assemblages are linked to the major boundary current off eastern Australia"
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [ASG]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - ASG: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species, probabilistically, and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used seabird at-sea ship-based survey data collected over 37° latitude and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different among them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic layers.
  **Results**: All models suggested two large scale assemblages ('northern' and 'southern'), except for the autumn presence-absence model which indicated three groups. Model results seem to agree in a biogeographic barrier at ~34°S, where the East Australian Current dissipates as it flows eastward and loses force. Moreover, sea surface temperature or sea surface salinity, were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Conclusion**: RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and large spatial scales. Given the rapid change that the East Australian Current is experiencing, and its importance in driving distributions of seabirds, the methods applied in our study could be replicated in future analyses to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
output: 
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

@@

Bioregionalization is the discipline that identifies, delimits, characterises, and names biogeographical regions [@vilhena2015; @morrone2018]. Although there is often no agreement on the naming systems between proposed regionalizations [but see @ebach2008], the hierarchical nature of bioregions is well recognized [@vilhena2015; @kreft2010]. Thus, bioregions are usually described within geographically larger classifications, such as 'provinces' or 'realms' [@spalding2007; @costello2017]. The delimitation of bioregions is made through identifying sets of species that are endemic or have high levels of co-occurrence, and are distinct enough from other regions. In this sense, bioregions are interpreted as a natural representations of common historical and/or ecological processes shaped by physical and biological forces [@kreft2010]. In particular, the identification of bioregions allows representation of the distribution of multiple species at a given time and space as well as relating these regions to environmental characteristics. By doing so, bioregions can be interpreted as species assemblages [@lyons2017; @hill2017]. Given these features, bioregions can be used as surrogates for biodiversity monitoring in data-poor regions and in regional conservation planning [@spalding2007; @ferrier2002a], playing a fundamental role in conservation biogeography [@whittaker2005].

Seabird assemblages have been widely described and related to environmental characteristics. Considering the hierarchical nature of oceanographic processes and their resulting features [@haury1978patterns], it is not surprising that the distribution of highly mobile seabirds are linked differently across scales [@hunt1987]. In the marine environment, large scale oceanographic features (e.g. water masses and currents) act as 'biomes' and influence seabird assemblage structure--i.e. the occurrence of species and their relative numerical composition [@ribic1997; @hyrenbach2007; @commins2014]. In contrast, local abundance is related to temporary processes that aggregate prey and therefore predators, such as fronts and eddies [@weimerskirch2007; @scales2014]. Such smaller-scale processes, although ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur [@davoren2013; @hazen2013]. Noticeably, sea surface temperature is a key driver of seabird distribution, at individual- and assemblage-levels, operating over a range of scales from large (e.g. water masses) to small (e.g. fronts) [@dunlop1988; @ribic1997; @quillfeldt2015]. Nonetheless, many regions still lack information regarding distribution of seabird species and how they assemble in space and time.

One area that could benefit from better understanding of seabird distributions is Australia, considered a world seabird hotspot [@karpouzi2007; @ramírez2017]. Given the region's importance for seabird diversity, there are surprisingly few studies on seabird distribution [@mott2018; @bernard2021]. Off northwest Australia, at-sea surveys, multivariate analyses [@dunlop1988] and ensemble species distribution models of species assemblages [@lavers2014] have shown that habitat use by tropical species is related to sea surface temperature and salinity. In particular, @dunlop1988 suggested that salinity is the main variable influencing species distribution in oceanic waters whereas temperature is the key factor on the continental shelf. In contrast, there are few studies at large scales off eastern Australia. @reid2000 and @evans2021a analysed seabird assemblages and their relationships with oceanography and prey data at broad scales (tens of km), off southeast Tasmania. The remaining studies are mostly descriptive accounts at local scales [e.g. @norman1992distribution; @blaber1994distribution], or focused on a single species [e.g. @priddel2014].

The East Australian Current (EAC) is the major western boundary current in the South Pacific, and the dominant current in eastern Australia [@ganachaud2014]. The EAC has been extensively studied, although its link between oceanography and biological processes is still in its infancy. The dynamic features caused by the EAC, such as eddies and upwellings, are likely to influence marine biota [@suthers2011; @scales2014]. While most of the biological studies done in this system were focused on plankton entrapment within the cores of eddies [@garcia2022; @firme2023; @hassler2011], there is a growing body of work focusing on fisheries [e.g. @hobday2010; @young2011; @brieva2015]. These studies emphasize how the probability of occurrence of a given taxon, and the structure of species assemblages, is driven by the EAC. Notwithstanding, the relationship between oceanography and marine megafauna along the EAC is poorly known [@suthers2011]. Some progress has been made using shark and ray movement data [@jaine2014; @niella2020; @niella2022], and satellite tracking has also demonstrated that increased sea surface temperature in EAC decreases penguin foraging success [@carroll2016]. However, information at the assemblage-level is still scarce [but see @evans2021a].

Marine species range shifts are more likely to occur faster than terrestrial species due to their higher sensitivity to warming [@pinsky2019; @lenoir2020]. As a result, climate change is rearranging the structure of species assemblages around the globe [@poloczanska2013; @antão2020; @dornelas2014]. Top predators, for instance, may change their core habitats [@hazen2013a; @reisinger2022], with unprecedented population consequences. For seabirds, @péron2010 showed that sea surface temperature anomalies affected the occurrence and abundance of albatrosses and petrels species along a latitudinal range in the Indian Ocean, thus affecting assemblage structure. In eastern Australia, changing EAC conditions have affected a range of species. Changes in zooplankton assemblage structure and fish distribution patterns have been recorded, and shark distribution shifts are expected to occur [@johnson2011; @niella2022]. One of the strongest marine heatwave ever recorded, in the Tasman Sea, showed drastic environmental impacts [@oliver2017]. Therefore, given that the EAC is strengthening and getting warmer [@cai2005; @phillips2022], biological changes are expected to occur [@poloczanska2007; @hobday2010].

Given east Australia's importance for seabirds and the recent attention on how climate change in the EAC can affect marine biota [@poloczanska2007], an assessment of the spatial distribution of seabirds at large scales is timely. Seabirds can serve as sentinels for tracking ecological changes in the EAC in many ways [e.g. @sydeman2012]. Surprisingly, though, seabird distribution is under-represented in climate change studies, despite being highlighted as a priority topic [@grémillet2009; @sydeman2012]. In this study, we aim to identify assemblages of seabird species probabilistically and relate them to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We used the mixture model Region of Common Profiles [RCP; @foster2013], regarded as one of the best available tools for describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. East Australia supports a high diversity of seabirds and offers wide environmental gradients, from the tropics to the subantarctic. Therefore, we expect to find different assemblages relating to different environmental conditions. In addition, we run RCP models for presence-absence and abundance as input data, to test if the different numeric resolution would result in similar assemblage patterns. Furthermore, we provide a repeatable framework on which future studies can build to assess how changes in the EAC will influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical, nutrient-poor water. When the SEC's North Caledonian Jet hits the Queensland Plateau (\~18°S), it splits into a north branch running towards the Gulf of Papua and a south branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig. 1a**; @kessler2013; @ganachaud2014). **SST/SSS characteristics? ROB?** The EAC interacts with the topography of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC is mainly attached to the shelf break, but at \~32°S it spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; often termed the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig. 1a**), possibly carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard RV Investigator and RV Falkor during 15 voyages, between 2016--2021, over \~37° of latitude (**Fig. 1b**; **Table S1**). One observer continuously recorded seabirds within a 300-m forward quadrant (eye height **18-m, Eric?** a.s.l) throughout daylight hours while the ship was underway at a constant speed. The species, the number of birds per group, and behaviour were logged directly into a data entry system through a ship-based web server. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species) **(Eric: is there any ref for the protocol?)**. Birds following the boat were previously excluded from analyses.

@@

To explain and predict species distribution and biogeography, a thoughtful choice of covariates is important [@barbet-massin2014]. We used three physiographic variables and seven oceanographic variables, from which a further three climatic layers were calculated (**Table 1**). These variables are the most used in seabird-environment relationship studies [@tremblay2009], capturing processes at a range of scales from large-scale water-mass properties (e.g. sea surface temperature) to prey availability in mesoscale oceanographic features (e.g. fronts and eddies; @weimerskirch2007; @scales2014). Seabird records were matched with the best open-access, remote-sensed or interpolated data available, and distance from the coast was calculated in R software, version 4.2.0 [@r2022] (see details below).

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022], using packages 'plyr' 1.8.8 [@plyr], 'dplyr' 1.1.2 [@dplyr], 'tidyr' 1.3.0 [@tidyr], 'readr' 2.1.4 [@readr], 'tibble' 3.2.1 [@tibble], 'lubridate' 1.9.2 [@lubridate], 'stringr' 1.5.0 [@stringr], 'purrr' 1.0.1 [@purrr], 'ggplot2' 3.4.2 [@ggplot2], 'ggspatial' 1.1.7 [@ggspatial], 'patchwork' 1.1.2 [@patchwork], 'RColorBrewer' 1.1-3 [@rcolorbrewer], 'rnaturalearth' 0.3.2 [@rnaturalearth], 'sp' 1.6-0 [@sp1; @sp2], 'sf' 1.0-8 [@sf], 'mapview' 2.11.0 [@mapview], 'raster' 3.5-21 [@raster], 'terra' 1.6-7 [@terra], 'rerddap' 1.0.2 [@rerddap], 'rerddapXtracto' 1.1.4 [@rerddapXtracto], 'hadsstR' [@hadsstr], 'corrplot' 0.92 [@corrplot2021], 'iNEXT' 3.0.0 [@chao2014; @hsieh2016], and 'ecomix' 1.0.0 [@ecomix]. Code and the modelling dataset (see \@ref(modellingdatasets)) are available in **Daudt et al. 2023 (Zenodo repo)**.

@@

Spatial and temporal persistence of features such as eddy kinetic energy (eke) and thermal fronts (sea surface temperature gradient; 'sst_grad') may influence the movement and distribution of seabirds by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Therefore, before extracting environmental data for each seabird record, 'eke' and 'sst_grad' environmental layers were calculated. For both, 'eke' and 'sst_grad', we used their original cell spatial resolution (**Table 1**) and calculations were done layer-by-layer to match seabird data on the finest scale possible. Eddy kinetic energy was calculated as $0.5 (u'² + v'²)$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ (**ROB: Ref here?**). The spatial 'sst_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS² + WE²$ [@burrows2011]. In addition, we calculated climatic seasonal average layers for 'eke' and 'sst_grad' based on 10-years of data (2011--2021) as a proxy for persistent oceanographic features, as well as 'eke' standard deviation over the same period (named with the prefix 'clim\_' in **Table 1**). Seabed slope was calculated using 'raster::terrain()', and distance from the coast was calculated through 'sf::st_distance()' using Australia and Tasmania as the mainland (i.e. ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from 'raster', 'terra', and 'sf' packages.

### Modelling datasets {#modellingdatasets}

Seabirds are highly mobile animals, and many species migrate during non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], as well as in probability of occurrence and total numbers. Thus, we build separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. 1b**), by summing seabirds from the same species. Considering our aim to test if presence-absence and abundance (count) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. Furthermore, to avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons. Environmental data were scaled and centred before modelling.

@@

We used the mixture model Region of Common Profiles (RCP) to identify and describe seabird assemblages. RCP is a one-stage bioregionalization approach, as it identifies bioregions by specifying a statistical model that simultaneously relates response (multi-species data) and predictor (environmental data) [@woolley2020; @hill2020]. The model defines 'regions' of similar environmental spaces in which the probability of observing a set of species 'profiles' is constant but different from other regions [@foster2013; @foster2017]. As such, RCPs can be directly interpreted as assemblages, and their species profiles as the assemblage structure. RCP memberships are latent (unobserved) and vary as a function of environmental data, assuming sites within each RCP have constant species profiles. Given a set of environmental variables, the model also enables prediction of assemblage probabilities onto sites where no biotic data exist. Additionally, as a feature of model-based approaches, RCPs allow the opportunity to compare models through information criteria, present uncertainty around the results, and check for model assumptions through assessing distribution of residuals [@warton2015; @hill2020].

Before fitting models, covariates were checked for collinearity within each seasonal modelling dataset. If a pairwise collinearity of $\left|x\right| >0.7$ was found we removed one of the variables [@dormann2013], favouring direct measures over the 'climatic' layers we calculated (e.g. if 'sst_grad' and 'clim_sst_grad' were correlated, we chose 'sst_grad'). Thus, 'full models' differed between seasons as the retained covariates varied, although most covariates were common for all seasons.

>>>>

---
title: "Seabird assemblages are linked to the major western boundary current off eastern Australia"
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [ASG]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - ASG: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species and relate them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used 15 seabird at-sea ship-based survey data collected over 37° of latitude, from 2016 to 2021, and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and abundance). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within regions, but different amongst them. These groups also vary according to covariates, which in our case included oceanographic, climatic, and physiographic parameters.
  **Results**: Based on 142,646 seabirds recorded, from 80 species, all models suggested two macro-scale assemblages (‘northern’ and ‘southern’), except for the autumn presence-absence model that identified three groups. The model results consistently show a biogeographic transition at ~34°S, near the latitude at which the East Australian Current separates from the Australian continental slope. Sea surface temperatures or sea surface salinities were selected in all final models, further indicating a close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and abundance, resulted in similar spatial and species profile patterns.
  **Conclusion**: RCP models clearly identified two seabird assemblages off the east coast of Australia, suggesting the persistence of these groups at seasonal and macro spatial scales. Given the rapid climatic changes that the East Australian Current is experiencing and its importance in influencing distributions of seabirds, the methods applied in our study could be replicated to assess possible changes in seabird assemblages, and how they are affected by changing environmental conditions.
output: 
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

@@

Bioregionalization is the discipline that identifies, delimits, characterises, and names biogeographical regions [@vilhena2015; @morrone2018]. Although there is often no agreement on the naming systems among proposed regionalization schema [but see @ebach2008], the hierarchical nature of bioregions is well recognized [@vilhena2015; @kreft2010]. Thus, bioregions are usually described within geographically larger classifications, such as 'provinces' or 'realms' [@spalding2007; @costello2017]. The delimitation of bioregions is made through identifying sets of species that are endemic or have high levels of co-occurrence, and are distinct enough from other regions. In this sense, bioregions are interpreted as a natural representations of common historical and/or ecological processes shaped by physical and biological forces [@kreft2010]. In particular, the identification of bioregions allows representation of the distribution of multiple species at a given time and space, and relating these regions to environmental characteristics. By doing so, bioregions can be interpreted as species assemblages [@lyons2017; @hill2017]. Given these features, bioregions can be used as surrogates for biodiversity monitoring in data-poor regions and in regional conservation planning [@spalding2007; @ferrier2002a], playing a fundamental role in conservation biogeography [@whittaker2005].

Seabird assemblages have been widely described and related to environmental characteristics [@ribic1997; @woehler2003; @hyrenbach2007]. Considering the hierarchical nature of oceanographic processes and their resulting features [@haury1978patterns], it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales [@haury1978patterns; @hunt1987]. In the marine environment, large scale (\>100 km length scale) oceanographic features (e.g. boundary currents and water masses) act as terrestrial biomes (based on vegetation), and influence seabird assemblage structure--i.e. the occurrence of species and their relative numerical composition [@ribic1997; @hyrenbach2007; @commins2014]. In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies [@weimerskirch2007; @scales2014]. Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur [@davoren2013; @hazen2013]. Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g. water masses) to fine (e.g. fronts) [@dunlop1988; @ribic1997; @quillfeldt2015]. Despite extensive surveys [e.g. @ballance1997] and tracking studies [e.g. @carneiro2020], many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

One area in need of better understanding of seabird distributions is Australia, considered a global seabird hotspot [@karpouzi2007; @ramírez2017]. Given the region's importance for seabird diversity, there are surprisingly few studies on seabird distributions at sea [@mott2018; @bernard2021]. Off northwest Australia, at-sea surveys, multivariate analyses [@dunlop1988] and ensemble species distribution models of species assemblages [@lavers2014] showed that habitat use by tropical seabird species was related to sea surface temperatures and salinity. @dunlop1988 suggested that salinity was the main variable influencing species' distributions in oceanic waters while temperature was the key factor for species on the continental shelf. In contrast, there are few studies at large spatial and temporal scales off eastern Australia. @reid2000 and @evans2021a analysed seabird assemblages and their relationships with oceanographic and prey data at coarse scales (tens of km) off southeast Tasmania. The remaining studies are mostly descriptive accounts locally [@norman1992distribution; @blaber1994distribution], or focused on a single species [@priddel2014].

The East Australian Current (EAC) is the major western boundary current in the South Pacific Ocean, and the dominant boundary current off eastern Australia [@ganachaud2014]. The EAC has been extensively studied, although research into the links between the EAC's oceanography and marine predators are not well known. The dynamic features associated with the EAC, such as mesoscale eddies and upwelling, are likely to influence marine biota [@suthers2011; @scales2014]. While most of the biological studies in this system were focused on plankton entrapment within the cores of eddies [@garcia2022; @firme2023; @hassler2011], there is a growing body of work focusing on fisheries [@hobday2010; @young2011; @brieva2015]. These studies emphasized how the probability of occurrence of a given taxon, and the structure of species assemblages, were driven by the EAC. However, the relationships between oceanography and marine megafauna within the EAC are poorly known [@suthers2011]. Some progress has been made using shark and ray movement data [@jaine2014; @niella2020; @niella2022], and satellite tracking has also demonstrated that increased sea surface temperatures in the EAC decreases little penguin *Eudyptula minor* foraging success [@carroll2016], but information at the assemblage-level for seabirds in the EAC is absent.

Marine species range shifts are more likely to occur faster than terrestrial species due to their higher sensitivity to warming [@pinsky2019; @lenoir2020]. As a result, climate change is rearranging the structure of species assemblages around the globe [@poloczanska2013; @antão2020; @dornelas2014]. Top predators, for instance, may change their core habitats [@hazen2013a; @reisinger2022], with unprecedented population consequences [@woehler2023]. For seabirds, @péron2010 and @sojitra2022 showed that sea surface temperature anomalies affected the occurrence and abundances of albatrosses and petrels species with latitude, thus affecting assemblage structure. In eastern Australia, changing EAC conditions have affected a range of marine species. For example, changes in zooplankton assemblage structure and fish distribution patterns have been recorded and attributed to warming water temperatures, and shifts in shark distributions are expected [@johnson2011; @niella2022]. A severe marine heatwave driven by a poleward strengthening of the EAC, showed drastic environmental impacts including disease outbreaks in aquaculture farms and wildlife mortalities [@oliver2017]. Given that the EAC is strengthening and extending farther southward and getting warmer [@cai2005; @phillips2022], biological changes are expected to occur [@poloczanska2007; @hobday2010].

In light of eastern Australia's importance for seabirds, and the recent attention on how predicted climate change will influence the EAC's marine biota [@poloczanska2007], an assessment of the contemporary spatial distributions of seabirds within the EAC at large temporal and spatial scales is timely. Seabirds can serve as sentinels for tracking ecological changes in the EAC in many ways [e.g. @sydeman2012]. Surprisingly though, seabird distribution is under-represented in climate change studies, despite being highlighted as a priority topic [@grémillet2009; @sydeman2012].

In this study, we aim to identify assemblages of seabird species within the EAC and adjacent areas, and relate these assemblages to environmental conditions, by using at-sea surveys data and a state-of-the-art modelling technique. We use the Region of Common Profiles [RCP; @foster2013], a mixture model regarded as one of the best available tools for describing bioregions [@hill2020; @woolley2020]. RCP is a model-based approach, and has a range of advantages over traditional algorithm-based statistics [@warton2015; @hill2020]. Eastern Australia supports a high diversity of seabirds and offers wide environmental gradients over a near 40° latitudinal range, from the tropics to the cool temperate (**ref?**) We hypothesised to find different seabird assemblages relating to different environmental conditions within the EAC. In addition, presence-absence and abundance data were used to test if the different numerical resolutions would result in the identification of similar assemblages. This study provides a repeatable framework on which future studies can compare to assess how future changes in the EAC may influence the distribution of seabird assemblages.

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig. 1a**; @kessler2013; @ganachaud2014). **SST/SSS characteristics? ROB?** The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flow polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension' (@oke2019a; often termed the 'Tasman Front'). Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig. 1a**), carrying eddy-like features as a transient current [@oke2019].

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Fig. 1b**; **Table S1**). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. **Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals.** Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

@@

To investigate and predict species' distributions and relationships with environmental variables, a considered choice of **broad-scale climatic and oceanographic** covariates is important [@barbet-massin2014]. We used three physiographic variables and seven oceanographic variables, from which a further three climatic layers were calculated (**Table 1**). These variables are the most frequently used in seabird-environment relationship studies [@tremblay2009], capturing processes at a range of spatial and temporal scales from macro-scale water-mass properties to prey availability in meso- to submesoscale oceanographic features [@haury1978patterns; @weimerskirch2007; @scales2014].

@@

Data wrangling, modelling and visualization were done in R 4.2.0 [@r2022]; the full list of packages can be found in the Supplementary Material **XX**. Code and the modelling dataset (see \@ref(modellingdatasets)) are available in **Daudt et al. 2023 (Zenodo repo)**.

<!-- using packages 'plyr' 1.8.8 [@plyr], 'dplyr' 1.1.2 [@dplyr], 'tidyr' 1.3.0 [@tidyr], 'readr' 2.1.4 [@readr], 'tibble' 3.2.1 [@tibble], 'lubridate' 1.9.2 [@lubridate], 'stringr' 1.5.0 [@stringr], 'purrr' 1.0.1 [@purrr], 'ggplot2' 3.4.2 [@ggplot2], 'ggspatial' 1.1.7 [@ggspatial], 'patchwork' 1.1.2 [@patchwork], 'RColorBrewer' 1.1-3 [@rcolorbrewer], 'rnaturalearth' 0.3.2 [@rnaturalearth], 'sp' 1.6-0 [@sp1; @sp2], 'sf' 1.0-8 [@sf], 'mapview' 2.11.0 [@mapview], 'raster' 3.5-21 [@raster], 'terra' 1.6-7 [@terra], 'rerddap' 1.0.2 [@rerddap], 'rerddapXtracto' 1.1.4 [@rerddapXtracto], 'hadsstR' [@hadsstr], 'corrplot' 0.92 [@corrplot2021]. -->

@@

The spatial and temporal persistence of oceanographic features such as eddy kinetic energy ('eke') and thermal fronts (sea surface temperature gradient: 'sst_grad') may influence the movements and distributions of seabirds by offering a higher likelihood of finding food [@weimerskirch2007; @scales2014a]. Thus, 'eke' and 'sst_grad' environmental layers were calculated before extracting environmental data for each seabird record. For both 'eke' and 'sst_grad', we used their original cell spatial resolution (**Table 1**) and calculations were done layer-by-layer to match seabird data on the finest spatial scale possible. Eddy kinetic energy was calculated as $0.5 (u'² + v'²)$, where $u' = u - \bar{u}$ and $v' = v - \bar{v}$ are the surface zonal (latitudinal) and meridian (longitudinal) geostrophic currents, respectively, using a rolling mean of 90 days as $\bar{u}$ and $\bar{v}$ [@sun2022]. The spatial 'sst_grad' for each cell was calculated by averaging the maximum value for each neighbouring cell (average neighbourhood maximum), from north-to-south (NS) and west-to-east (WE); then, the magnitude of the gradient was defined as $NS² + WE²$ [@burrows2011]. In addition, we calculated climatic seasonal average layers for 'eke' and 'sst_grad' based on the 11-year data for 2011--2021 as a proxy for persistent oceanographic features, and the 'eke' standard deviation over the same period (named with the prefix 'clim\_' in **Table 1**). Seabed slope was calculated using 'raster::terrain()' [@raster], and distance from the coast was calculated through 'sf::st_distance()' [@sf] using Australia and Tasmania as the mainland (i.e. ignoring smaller islands). Then, for each seabird record, the most proximate value for each environmental layer (original or calculated) was extracted using built-in functions from 'raster' 3.5-21, 'terra' 1.6-7, and 'sf' 1.0-8 packages [@sf; @raster; @terra].

### Data {#modellingdatasets}

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling dataset for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (c. 110 $\times$ 110 km) and season (**Fig. 1b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were **standardised** before modelling.

@@

We used the mixture model Region of Common Profiles (RCP) to identify and describe seabird species assemblages. RCP is a one-stage bioregionalization approach, as it identifies bioregions by specifying a statistical model that simultaneously relates response (multi-species data) and predictor (environmental data) [@woolley2020; @hill2020]. Given a set of environmental variables, the model can predict assemblage probabilities of sites where no biotic (observational) data exist. As a model-based approach, RCPs can be compared by information criteria, present uncertainty around the results, and check for model assumptions through assessing distribution of residuals [@warton2015; @hill2020].

The model is specified in terms of assemblages and site-specific probabilities. There are $K$ assemblages (i.e. RCPs) that are common across the study area, where $K = (1, ..., k)$. Depending on whether the model is specified in terms of abundance or presence-absence, the assemblages are defined in terms of their structure at each $i$th site (a vector $y_{i}$), i.e. frequency or relative frequency of the species, respectively. The assemblage structure is denoted by $y_{i} = (y_{i1}, ... y_{iS})$, where $S$ is the number of species. The model estimates the probability of belonging to the $K$ assemblage given the site's assemblage structure. Thus, each site within the study area has a $k$-dimensional probability vector that gives the probability of the corresponding assemblage for that site. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) [@foster2013]. However, the final estimated assemblage structure is assumed to be constant for all sites belonging to the $K$ assemblage [@foster2013]. As such, RCPs can be directly interpreted as assemblages, and their species profiles as the assemblage structure.

Before fitting models, covariates were checked for collinearity within each seasonal dataset. If a pairwise collinearity of $\left|x\right| >0.7$ was found we removed one of the variables [@dormann2013], favouring direct measures over the 'climatic' layers we calculated (e.g. if 'sst_grad' and 'clim_sst_grad' were correlated, we chose 'sst_grad'). Thus, 'full models' differed between seasons as the retained covariates varied, although most covariates were common for all seasons.

Due to seabird migration and dispersion, seasons could act as 'sampling artifacts' [@foster2017], thus biasing results [@ponti2023]. In light of the difference among seasons in the numbers of species that met our occurrence threshold (see \@ref(modellingdatasets)), we decided to use RCP models by season and data type (presence-absence and abundance). RCP models were used to relate the response variables (i.e. presence-absence or abundance of each seabird species in each grid cell) to the independent predictor covariates (i.e. averaged values of environmental variables in each cell). Presence-absence data were modelled based on a Bernoulli model, whereas abundance data based on a Negative Binomial model. RCPs were fitted using the 'ecomix' 1.0.0 package [@ecomix], with the following workflow: (i) find the best number of groups ($k$) that describes the data, (ii) re-fit the model by fixing $k$ and selecting optimal combination of covariates, (iii) check residuals, (iv) bootstrap the best model to estimate uncertainty around parameter estimates, and (v) produce results.
